---
title: "Employee Satisfaction Survey Analysis"
author: "Ibrahim Adippe"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)

library(tidyverse)
library(scales)
library(knitr)
library(kableExtra)

# Load data
survey <- read.csv("employee_survey.csv", stringsAsFactors = FALSE)
```

# Executive Summary

This report analyzes employee satisfaction data from a company-wide survey of **`r nrow(survey)` employees** across `r length(unique(survey$department))` departments.

```{r summary_stats, echo=FALSE}
avg_satisfaction <- mean(survey$q1_satisfaction, na.rm = TRUE)
pct_would_refer <- mean(survey$would_refer_friend == "Yes", na.rm = TRUE) * 100
lowest_score_area <- c(
  "Job Satisfaction" = mean(survey$q1_satisfaction, na.rm = TRUE),
  "Work-Life Balance" = mean(survey$q2_worklife_balance, na.rm = TRUE),
  "Manager Support" = mean(survey$q3_manager_support, na.rm = TRUE),
  "Career Growth" = mean(survey$q4_career_growth, na.rm = TRUE),
  "Would Recommend" = mean(survey$q5_recommend, na.rm = TRUE)
)
weakest_area <- names(which.min(lowest_score_area))
```

- **Average job satisfaction:** `r round(avg_satisfaction, 2)` / 5.0
- **`r round(pct_would_refer, 1)`%** would refer a friend to work here
- **Weakest area:** `r weakest_area` (avg `r round(min(lowest_score_area), 2)`/5)

# 1. Demographics

```{r demographics, echo=FALSE}
dept_summary <- survey %>%
  count(department, name = "employees") %>%
  mutate(percentage = percent(employees / sum(employees))) %>%
  arrange(desc(employees))

kable(dept_summary, 
      col.names = c("Department", "Employees", "% of Total"),
      caption = "Distribution by Department") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r dept_chart}
survey %>%
  count(department) %>%
  mutate(department = fct_reorder(department, n)) %>%
  ggplot(aes(x = department, y = n, fill = department)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.2, size = 4) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Employees by Department",
       x = NULL, y = "Number of Employees") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

# 2. Satisfaction Scores by Category

```{r satisfaction_overview}
scores <- survey %>%
  summarise(
    `Job Satisfaction` = mean(q1_satisfaction, na.rm = TRUE),
    `Work-Life Balance` = mean(q2_worklife_balance, na.rm = TRUE),
    `Manager Support` = mean(q3_manager_support, na.rm = TRUE),
    `Career Growth` = mean(q4_career_growth, na.rm = TRUE),
    `Would Recommend` = mean(q5_recommend, na.rm = TRUE),
    `Salary Satisfaction` = mean(salary_satisfaction, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "category", values_to = "avg_score")

scores %>%
  mutate(category = fct_reorder(category, avg_score)) %>%
  ggplot(aes(x = category, y = avg_score, fill = avg_score)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = round(avg_score, 2)), hjust = -0.2, size = 4) +
  coord_flip() +
  scale_fill_gradient(low = "#E74C3C", high = "#27AE60") +
  scale_y_continuous(limits = c(0, 5), breaks = 0:5) +
  labs(title = "Average Scores by Category (1-5 Scale)",
       subtitle = "Higher = better. Color indicates performance.",
       x = NULL, y = "Average Score") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

# 3. Satisfaction by Department

```{r dept_satisfaction}
dept_scores <- survey %>%
  group_by(department) %>%
  summarise(
    `Job Satisfaction` = mean(q1_satisfaction, na.rm = TRUE),
    `Work-Life Balance` = mean(q2_worklife_balance, na.rm = TRUE),
    `Manager Support` = mean(q3_manager_support, na.rm = TRUE),
    `Career Growth` = mean(q4_career_growth, na.rm = TRUE)
  ) %>%
  pivot_longer(-department, names_to = "metric", values_to = "score")

ggplot(dept_scores, aes(x = department, y = metric, fill = score)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = round(score, 2)), color = "white", fontface = "bold") +
  scale_fill_gradient2(low = "#E74C3C", mid = "#F39C12", high = "#27AE60",
                       midpoint = 3, limits = c(1, 5)) +
  labs(title = "Satisfaction Heatmap: Department × Category",
       x = NULL, y = NULL, fill = "Score") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

# 4. Impact of Work Mode

Does where people work affect their satisfaction?

```{r work_mode_impact}
work_mode_stats <- survey %>%
  group_by(work_mode) %>%
  summarise(
    n = n(),
    avg_satisfaction = round(mean(q1_satisfaction, na.rm = TRUE), 2),
    avg_worklife = round(mean(q2_worklife_balance, na.rm = TRUE), 2),
    avg_career = round(mean(q4_career_growth, na.rm = TRUE), 2),
    pct_would_refer = round(mean(would_refer_friend == "Yes", na.rm = TRUE) * 100, 1)
  )

kable(work_mode_stats,
      col.names = c("Work Mode", "N", "Job Satisfaction", 
                    "Work-Life Balance", "Career Growth", "% Refer Friend"),
      caption = "Satisfaction Metrics by Work Mode") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

```{r work_mode_chart}
survey %>%
  select(work_mode, q1_satisfaction, q2_worklife_balance) %>%
  pivot_longer(-work_mode, names_to = "metric", values_to = "score") %>%
  mutate(metric = recode(metric,
                         q1_satisfaction = "Job Satisfaction",
                         q2_worklife_balance = "Work-Life Balance")) %>%
  ggplot(aes(x = work_mode, y = score, fill = work_mode)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.8) +
  facet_wrap(~metric) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Score Distribution by Work Mode",
       x = NULL, y = "Score (1-5)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold"))
```

# 5. Referral Intent

```{r nps}
refer_stats <- survey %>%
  count(would_refer_friend) %>%
  mutate(pct = n / sum(n) * 100,
         would_refer_friend = factor(would_refer_friend, 
                                     levels = c("Yes", "Maybe", "No")))

ggplot(refer_stats, aes(x = "", y = pct, fill = would_refer_friend)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("Yes" = "#27AE60", 
                               "Maybe" = "#F39C12", 
                               "No" = "#E74C3C")) +
  geom_text(aes(label = paste0(round(pct, 1), "%")), 
            position = position_stack(vjust = 0.5), 
            color = "white", fontface = "bold", size = 5) +
  labs(title = "Would You Refer a Friend to Work Here?",
       fill = "Response") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

# 6. Key Recommendations

```{r recommendations, echo=FALSE, results='asis'}
weakest <- scores %>% arrange(avg_score) %>% head(2) %>% pull(category)
strongest <- scores %>% arrange(desc(avg_score)) %>% head(2) %>% pull(category)

cat("**Priority areas for improvement:**\n\n")
cat(paste0("- ", weakest, "\n"), sep = "")
cat("\n**Areas performing well (maintain):**\n\n")
cat(paste0("- ", strongest, "\n"), sep = "")
```

# Methodology

- **Sample size:** `r nrow(survey)` employees
- **Scale:** 1 (Strongly Disagree) to 5 (Strongly Agree)
- **Missing data:** Handled with pairwise deletion (`na.rm = TRUE`)
- **Tools:** R `r paste(R.version$major, R.version$minor, sep=".")`, tidyverse, ggplot2

---

*Report generated on `r format(Sys.Date(), "%B %d, %Y")`*